---
title: "RepData assignment Lei"
author: "Lei"
date: "12 Juni 2017"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Project assignment of week 2, Coursera course "Replicative research".

### Load and preprocess data
```{r}
activity <- data.frame(read.csv("activity.csv", header = TRUE))
dim(activity)
head(activity)
```



### Question 1:
What is the mean total number of steps taken per day?

Process data for plotting:
```{r}
aggreSteps <- aggregate(activity$steps, list(activity$date), FUN = sum)
#change names for the aggregated table:
names(aggreSteps)[[1]] <- "Date"
names(aggreSteps)[[2]] <- "Steps"
# delete rows with NaN in steps:
aggreSteps <- aggreSteps[complete.cases(aggreSteps),]

```

Make histogram of the total number of steps taken each day:

```{r hist, echo=FALSE}
hist(aggreSteps$Steps, col = "green", main = "Number of steps per day",
     xlab = "Steps")

```

### Question 2:
Calculate and report the mean and median of total number of steps taken per day:
```{r mean}
steps_mean <- mean(aggreSteps$Steps)
```

```{r median}
steps_median <- median(aggreSteps$Steps)
```

The mean value of the total number of steps taken per day is `r steps_mean`. 
The median of the the total number of steps taken per day is `r steps_median`.


### What is the average daily activity pattern?
Make a time series plot (line chart) of the 5-minute interval and the average number of steps taken, averaged across all days:

- first get average steps for each interval --> aggregate
- then plot steps (as y) against interval
- use line chart (type="l")


```{r aggregate}
step_by_inteval <- aggregate(steps ~ interval, activity, mean)
plot(step_by_inteval$interval, step_by_inteval$steps, type = "l",
     main="Mean of steps per day over interval",
     xlab = "Interval", ylab="Steps", col="red")
```

```{r get max}
max_interval <- step_by_inteval$interval[which.max(step_by_inteval$steps)]

```

The 5-minute interval `r max_interval` corresponds with the maximum number of steps.

### Imputing missing values
Substitute the missing values (NAs) in the dataset with pertinent strategy.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).


```{r}
NA_rows <- nrow(activity) - sum(complete.cases(activity))

```

There are `r NA_rows` incomplete rows in the dataset.

2. Filling in the missing values in the dataset with median for that day:
(some how I have not made it with apply...)

```{r}
mean_per_day <- aggregate(steps~date, activity, FUN = mean, na.rm=TRUE)
names(mean_per_day)[[2]] <- "Mean_steps"

# try lookup:
filled_activity <- activity #copy activity to a new df
# lookup and replace NAs:
#filled_activity[] <- mean_per_day$Mean_steps[match(unlist(activity), mean_per_day$date)]

# for loop wroks at least:
for(i in 1:nrow(activity) ){
    if (is.na(activity[i,"steps"])){
        activity[i,"steps"] <- mean_per_day[activity[i,"date"],"Mean_steps"]
    }
}

head(filled_activity)
summary(filled_activity)
```

Again, make histogram of the total number of steps taken each day, this time with filled dataset:

```{r}
aggreSteps_filled <- aggregate(filled_activity$steps, list(filled_activity$date), FUN = sum)
#change names for the aggregated table:
names(aggreSteps_filled)[[1]] <- "Date"
names(aggreSteps_filled)[[2]] <- "Steps"

```

```{r hist_filled, echo=FALSE}
hist(aggreSteps_filled$Steps, col = "blue", main = "Number of steps per day",
     xlab = "Steps")

```

Again, calculate and report the mean and median of total number of steps taken per day, here for filled datasets (without NAs):
```{r mean_filled}
steps_mean_filled <- mean(aggreSteps_filled$Steps, na.rm = TRUE)
```

```{r median_filled}
steps_median_filled <- median(aggreSteps_filled$Steps, na.rm = TRUE)
```

The mean value of the total number of steps taken per day is `r steps_mean_filled`. 
The median of the the total number of steps taken per day is `r steps_median_filled`.

It seems that the NA substitution did not change the mean and median of the dataset, which is good.

### Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r week}
library(dplyr)
filled_activity_w <- mutate(filled_activity, weekend = weekdays(as.Date(filled_activity$date)) %in% c('Samstag','Sonntag')) ## Note: German version of RStudio!

```

2. Make lattice plot for weekends (TRUE in the weekend column) and for weekdays:

```{r lattice plot}
library(lattice)
# create factors with value labels
filled_activity_w$weekend.f<-factor(filled_activity_w$weekend,levels=c("FALSE","TRUE"),
   labels=c("Weekdays","Weekends")) 

# second try with helping hand:
steps_by_interval_filled <- aggregate(steps ~ interval + weekend.f, filled_activity_w, mean)

xyplot(steps ~ interval| weekend.f, 
       data = steps_by_interval_filled,
       main="Compare weekday and weekdays", 
       ylab="Number of steps",layout=c(1,2), type="l")

```


